---
title: "Analysis"
author: "Tsunghan Hsieh"
date: "2022-08-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

This analysis is aiming for identifying aging-associated genes (biomarkers) in large intestine and visualizing it in Tableau. The data is subset from [Tabula muris](https://s3.console.aws.amazon.com/s3/buckets/czb-tabula-muris-senis?region=us-west-2&tab=objects). The subset data is then filtered and normalized by [my python script](https://github.com/pocession/Tabula_muris/scRNA_subsetting.ipynb).

# Files

Due to the large size, all input files are not uploaded. All output files for Tableau are saved in [output folder](https://github.com/pocession/Tabula_muris/output).

# Configure

## Packages

Packages are required for this analysis.

```{r packages, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(Matrix)
library(DESeq2)
library(msigdbr)
library(forcats)
library(fgsea)
```

## Set dir and variables

```{r dir, echo=FALSE}
wd <- getwd()
subset <- "subset"
output <- "output"
input_dir <- file.path(wd,subset)
output_dir <- file.path(wd,output)
method <- "facs"
tissue <- "LargeIntestine"
```

## functions

Functions used in this script

```{r functions, echo=FALSE}
## get pseudobulk for each cell type
getPseudobulk <- function(mat, celltype) {
  stopifnot(!missing(mat) || !missing(celltype))
   mat.summary <- do.call(cbind, lapply(levels(celltype), function(ct) {
     cells <- names(celltype)[celltype==ct]
     pseudobulk <- rowSums(mat[, cells])
     return(pseudobulk)
   }))
   colnames(mat.summary) <- levels(celltype)
   return(mat.summary)
}

## save to csv
save_csv <- function(df,dir,name) {
  stopifnot(!missing(df) || !missing(dir) || !missing(name))
  write.csv(df,file.path(dir,name))
}

## calculate row-wise variance 
RowVar <- function(x, ...) {
  rowSums((x - rowMeans(x, ...))^2, ...)/(dim(x)[2] - 1)
}

## check the differentially-expressed genes
checkDEG <- function(df,padj,log2FC) {
  stopifnot(!missing(df) || !missing(padj) || !missing(log2FC))
  df <- df %>%
    mutate(DEG = ifelse(padj > {{padj}}, "None", ifelse(
      log2FoldChange > {{log2FC}}, "Up", ifelse(
        log2FoldChange < -{{log2FC}}, "Down", "None"
      )
    )))
  df$symbol <- rownames(df)
  return(df)
}

## Get GO objects
get_GO <- function(df,pathway){
  stopifnot(!missing(df))
  stopifnot(!missing(pathway))
  input <- df
  input <- input[!is.na(input$log2FoldChange), ]
  input <- input %>%
    dplyr::arrange(desc(log2FoldChange))
  ranks <- setNames(input$log2FoldChange , input$symbol)
  fgseaRes <- fgsea(pathways = pathways, stats = ranks, minSize=15, maxSize=500)
  fgseaRes$topGenes <- ""
  for(k in seq(1:length(fgseaRes$leadingEdge))){
    fgseaRes$topGenes[k] <- fgseaRes$leadingEdge[k][[1]] %>% paste(., collapse = ",")
  }
  fgseaRes <- as.data.frame(fgseaRes)
  return(fgseaRes)
}

## Get GSEA plot
get_fgsea_plot <- function(df){
  stopifnot(!missing(df))
  df <- df %>%
    dplyr::mutate(Regulation = ifelse(NES > 0, "Up", "Down")) %>%
    dplyr::mutate(log_padj_mutated = ifelse(NES > 0, -log10(padj), log10(padj)))
  df_up <- df %>%
    dplyr::arrange(desc(NES)) %>%
    slice(1:5)
  df_down <- df %>%
    dplyr::arrange(NES) %>%
    slice(1:5)
  df_subset <- rbind(df_up,df_down)
  
  p.fgsea <- ggplot(df_subset,aes(x=reorder(pathway,log_padj_mutated),y=log_padj_mutated,fill=factor(Regulation))) + 
    geom_bar(stat='identity') + theme(panel.background = element_rect(fill = "white", colour = "black"),
                                      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                      plot.title = element_text(size = 7),
                                      axis.title.y = element_blank(), axis.title.x = element_blank(),
                                      axis.text= element_text(size = 5)) + 
    geom_hline(yintercept = -log10(0.05)) + 
    geom_hline(yintercept = log10(0.05)) + 
    scale_fill_manual(values=c("blue","red")) +
    coord_flip() + labs(y='-log10(padj)',x='pathways')
  return(p.fgsea)
}
```

## read inputs

```{r inputs, echo=FALSE}
annot <- read.csv(file.path(input_dir,paste(method,tissue,paste0("annot",".csv"),sep="_")))
mtx <- read.csv(file.path(input_dir,paste(method,tissue,paste0("mtx",".csv"),sep="_"))) # long process
rownames(mtx) <- mtx$index
mtx <- mtx[,2:ncol(mtx)]
mtx <- as.matrix(t(mtx))

## convert count matrix to sparse matrix for faster computing
mtx.sparse <- Matrix(mtx, sparse = TRUE)

## Check the sparse matrix class
class(mtx.sparse)
```

# Analysis process

## Cell proportion dynamics

Generate reports for cell proportion dynamics

```{r cell_proportions, warnings=FALSE, message=FALSE}
type_cell_proportion <- "cell_proportion"
cell_sum <- annot %>%
  group_by(mouse.id,cell_ontology_class) %>%
  summarise(count=n()) %>%
  summarise(sum=sum(count))
cell_proportion <- annot %>%
  group_by(mouse.id,cell_ontology_class) %>% 
  summarise(count=n()) %>% 
  left_join(cell_sum,by="mouse.id") %>%
  separate(mouse.id,c("month","id","sex"),sep="_") %>%
  mutate(proportion = 100*(count/sum))
save_csv(cell_proportion,output_dir,paste(method,tissue,paste0(type_cell_proportion,".csv"),sep="_"))
```

## Expression dynamics in each cell types

Take a look at how many cell types in this tissue.

```{r cell_proportions, warnings=FALSE}
cell.type.list <- annot %>%
  group_by(cell_ontology_class) %>%
  summarise(count=n())

## Determine the major and minor cell population
minor_cell_populaiton <- c("Brush cell of epithelium proper of large intestine",
                           "secretory cell",
                           "enteroendocrine cell")
major_cell_population <- c("all",
                           "intestinal crypt stem cell",
                           "large intestine goblet cell",
                           "enterocyte of epithelium of large intestine",
                           "epithelial cell of large intestine")
cell.type.list
```

Take a look at how many mouse id.

```{r cell_proportions, warnings=FALSE}
mouseid.list <- annot %>%
  distinct(mouse.id) %>%
  as.list(.)
mouseid.list
```

### Get data with the most variant genes

To reduce the computation, we will only choose top genes with the most variance for further analysis.

```{r variance, warnings=FALSE}
## Calculate the variance, sort the gene based on the variance, and get the top variant genes
Gene.Var <- RowVar((mtx.sparse)) %>%
  sort(decreasing = TRUE)
Gene.subset <- Gene.Var[1:22075]

## Subset data based on the gene index
mtx.sparse.subset <- mtx.sparse[names(Gene.subset),]
```

### Get pseudobulk for each cell type in each mouse

```{r pseudobulk mouse, warnings=FALSE}
## remove minor cell populations, because in some mouse, there are only few cells from those population 
mouse.list <- c("18_53_M", "18_45_M", "18_46_F",
                 "18_47_F", "24_60_M", "24_59_M",
                 "24_58_M", "24_61_M", "3_8_M",
                 "3_9_M", "3_10_M", "3_11_M",
                 "3_56_F", "3_38_F", "3_39_F" )
annot.subset <- annot %>%
  filter(!cell_ontology_class %in%  minor_cell_populaiton)

pseudobulk.mouse <- list()
for (imouse in mouse.list) {
  test.mouse <- annot.subset %>%
      filter(mouse.id == {{imouse}})
  celltype <- as.factor(test.mouse$cell_ontology_class)
  names(celltype) <- test.mouse$index
  output <- getPseudobulk(mtx.sparse.subset, celltype)
  output.df <- as.data.frame(output)
  output.df$all <- rowSums(output.df)
  pseudobulk.mouse[[imouse]] <- output.df
}
```

### For each cell type, combine the pseudobulk data from each mouse

```{r final pseudobulk, warning=FALSE}
pseudobulk.cell <- list()
for (icell in major_cell_population) {
  output <- data.frame(symbol=rownames(mtx.sparse.subset))
  for (imouse in mouse.list) {
    tmpdf <- data.frame(tmp = pseudobulk.mouse[[imouse]][,icell])
    colnames(tmpdf) <- imouse
    output <- cbind(output,tmpdf)
  }
  pseudobulk.cell[[icell]] <- output
}

for (icell in major_cell_population) {
  save_csv(pseudobulk.cell[[icell]],output_dir,
           paste(method,tissue,paste0(icell,"pseudobulk",".csv"),sep="_"))
}
```

## Differential analysis

Get differential genes among ages for each cell types

### Make col data for DeSeq2

Get meta data from annotation files and make it as `colData` object for use in `DESeq`.

```{r meta}
meta <- annot %>%
  select(mouse.id) %>%
  mutate(id = mouse.id) %>%
  distinct %>%
  separate(id,c("month","id","gender"))
rownames(meta) <- meta$mouse.id
meta$month.factor <- factor(meta$month)
meta$gender.factor <- factor(meta$gender)
```

### Read the count matrix

Read the count matrix by `DESeqDataSetFromMatrix()`.

```{r read_count, message=FALSE}
dds.list <- list()
for (icell in major_cell_population) {
  input <- pseudobulk.cell[[icell]]
  test.df <- input[,2:ncol(input)] %>%
    mutate_if(is.numeric,round)
  rownames(test.df) <- input$symbol
  test.mtx <- as.matrix(test.df)
  dds.list[[icell]] <- DESeqDataSetFromMatrix(countData = test.mtx,
                                colData = meta,
                                design = ~ month.factor)
}
```

### Perform differential analysis (18 vs 3 and 24 vs 3)

Perform differential analysis by comparing 18 vs 3 months and 24 vs 3 months. Also get the count matrix (We already normalized it before, so we don't normalize it again) 

```{r differential_analysis, message=FALSE}
res.list <- list()
for (icell in major_cell_population) {
  dds <- DESeq(dds.list[[icell]])
  for (imonth in c("18","24")) {
    res <- as.data.frame(results(dds,contrast = c("month.factor",imonth,"3")))
    res <- checkDEG(res, 0.05, 0.5)
    res.list[[icell]][[imonth]] <- res
  }
}

## combine all counts
count.list <- list()
for (icell in major_cell_population) {
  input <- dds.list[[icell]]
  input <- estimateSizeFactors(input)
  count.list[[icell]] <- as.data.frame(counts(input, normalized=TRUE))
}
```

### Generate data frame for Tableau and save it

```{r Tableau}
df.list <- list()
for (icell in major_cell_population) {
  df <- as.data.frame(t(count.list[[icell]]))
  df$mouse.id <- rownames(df)
  df <- df %>%
    gather(symbol, count, -(mouse.id)) %>%
    separate(mouse.id, c("month","id","gender")) %>%
    mutate(mouse.id = paste(month,id,gender,sep="_")) %>%
    left_join(res.list[[icell]][["18"]][,c("symbol","log2FoldChange","DEG")],by="symbol") %>%
    left_join(res.list[[icell]][["24"]][,c("symbol","log2FoldChange","DEG")],by="symbol") %>%
    mutate(DEG.both = ifelse((DEG.x != "None") & (DEG.y != "None"), TRUE, FALSE))
  colnames(df) <- c("month","id","gender","symbol","count","mouse.id",
                    "log2FC.18","DEG.18","log2FC.24","DEG.24","DEG.both")
  save_csv(df,output_dir,paste(method,tissue,icell,paste0("final_df",".csv"),sep="_"))
  df.list[[icell]] <- df
}
```


## Perform GSEA (Gene Set Enrichment analysis)

We perform GSEA for data from differential analysis. Keep in mind we may not have significant results since the number of differential genes in each cell type are low.

### Loading pathway

Load pathways from database.

```{r pathway, warning=FALSE, message=FALSE, results='hold'}
# GO_collections <- msigdbr_collections() 
# split() comes from base R
c2cpKegg_gene_sets = msigdbr(species = "Mus musculus", category = "C2", subcategory = "CP:KEGG")
c2cpKegg_gene_sets_list = split(x = c2cpKegg_gene_sets$gene_symbol, f = c2cpKegg_gene_sets$gs_name)

c2cpReactome_gene_sets = msigdbr(species = "Mus musculus", category = "C2", subcategory = "CP:REACTOME")
c2cpReactome_gene_sets_list = split(x = c2cpReactome_gene_sets$gene_symbol, f = c2cpReactome_gene_sets$gs_name)

c5bp_gene_sets = msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:BP")
c5bp_gene_sets_list = split(x = c5bp_gene_sets$gene_symbol, f = c5bp_gene_sets$gs_name)

# c5mf_gene_sets = msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:MF")
# c5mf_gene_sets_list = split(x = c5mf_gene_sets$gene_symbol, f = c5mf_gene_sets$gs_name)

gene_sets_list <- c(c2cpKegg_gene_sets_list, c2cpReactome_gene_sets_list, c5bp_gene_sets_list)

# Remove duplicated genes in the list
for(i in 1:length(gene_sets_list)){
  gene_sets_list[[i]] <- unique(gene_sets_list[[i]])
}
```

### Perform GSEA
```{r DEG_list, warning=FALSE, message=FALSE}
DEG.list <- list()
for (icell in major_cell_population) {
  DEG.list[[icell]] <- df.list[[icell]] %>%
    select(symbol,log2FC.24) %>%
    distinct(symbol,log2FC.24) %>%
    mutate(log2FoldChange = log2FC.24) %>%
    arrange(desc(log2FoldChange))
}

## Perform GSEA on only psudobulk data
pathways <- gene_sets_list
GO.list <- list()
for (icell in major_cell_population) {
  GO.list[[icell]] <- get_GO(DEG.list[[icell]],pathways)
}
```

### GSEA

Now we perform GSEA and save the data for Tableau.

```{r warning=FALSE, message=FALSE, results='hold'}
pathways <- gene_sets_list
GO.list <- list()
for (icell in major_cell_population) {
  GO.list[[icell]] <- get_GO(DEG.list[[icell]],pathway)
}

for (icell in major_cell_population) {
  output_df <- apply(GO.list[[icell]],2,as.character)
  save_csv(output_df,output_dir, paste(method,tissue,icell,paste0("pathway",".csv"),sep="_"))
}
```
